---
title: "Functional programming"
---

## Functionals

```{r}
randomise <- function(f) f(runif(1e3))
randomise(mean)
#> [1] 0.506
randomise(mean)
#> [1] 0.501
randomise(sum)
#> [1] 489
```

```{r}
library(purrr)
```

The most fundamental functional is purrr::map(). It takes a vector and a function, calls the function once for each element of the vector, and returns the results in a list. 

In other words, map(1:3, f) is equivalent to list(f(1), f(2), f(3)).

The implementation of map() is quite simple. We allocate a list the same length as the input, and then fill in the list with a for loop. The heart of the implementation is only a handful of lines of code:

```{r}
simple_map <- function(x, f, ...) {
  out <- vector("list", length(x))
  for (i in seq_along(x)) {
    out[[i]] <- f(x[[i]], ...)
  }
  out
}
```

The real purrr::map() function has a few differences: it is written in C

The base equivalent to map() is lapply(). The only difference is that lapply() does not support the helpers that you’ll learn about below, so if you’re only using map() from purrr, you can skip the additional dependency and use lapply() directly.

### Producing atomic vectors

map() returns a list, which makes it the most general of the map family because you can put anything in a list. But it is inconvenient to return a list when a simpler data structure would do, so there are four more specific variants: map_lgl(), map_int(), map_dbl(), and map_chr(). Each returns an atomic vector of the specified type:

```{r}
# map_chr() always returns a character vector
map_chr(mtcars, typeof)

# map_lgl() always returns a logical vector
map_lgl(mtcars, is.double)

# map_int() always returns a integer vector
n_unique <- function(x) length(unique(x))
map_int(mtcars, n_unique)

# map_dbl() always returns a double vector
map_dbl(mtcars, mean)
```

purrr uses the convention that suffixes, like _dbl(), refer to the output. All map_*() functions can take any type of vector as input. These examples rely on two facts: mtcars is a data frame, and data frames are lists containing vectors of the same length. This is more obvious if we draw a data frame with the same orientation as vector: (picture)

All map_* functions always return an output vector the same length as the input, which implies that each call to .f must return a single value. If it does not, you’ll get an error:

```{r}
pair <- function(x) c(x, x)
map_dbl(1:2, pair)
```

This is similar to the error you’ll get if .f returns the wrong type of result:

```{r}
map_dbl(1:2, as.character)
```

In either case, it’s often useful to switch back to map(), because map() can accept any type of output. That allows you to see the problematic output, and figure out what to do with it.

```{r}
map(1:2, pair)

map(1:2, as.character)
```

Base R has two apply functions that can return atomic vectors: sapply() and vapply(). I recommend that you avoid sapply() because it tries to simplify the result, so it can return a list, a vector, or a matrix. This makes it difficult to program with, and it should be avoided in non-interactive settings.

### Anonymous functions and shortcuts

Instead of using map() with an existing function, you can create an inline anonymous function:

```{r}
map_dbl(mtcars, function(x) length(unique(x)))
```

Anonymous functions are very useful, but the syntax is verbose. So purrr supports a special shortcut:

```{r}
map_dbl(mtcars, ~ length(unique(.x)))
```

This works because all purrr functions translate formulas, created by ~ (pronounced “twiddle”), into functions. 

This shortcut is particularly useful for generating random data:

```{r}
x <- map(1:3, ~ runif(2))
str(x)
```

Reserve this syntax for short and simple functions. A good rule of thumb is that if your function spans lines or uses {}, it’s time to give it a name.

The map functions also have shortcuts for extracting elements from a vector, powered by purrr::pluck(). You can use a character vector to select elements by name, an integer vector to select by position, or a list to select by both name and position. These are very useful for working with deeply nested lists, which often arise when working with JSON.

```{r}
x <- list(
  list(-1, x = 1, y = c(2), z = "a"),
  list(-2, x = 4, y = c(5, 6), z = "b"),
  list(-3, x = 8, y = c(9, 10, 11))
)

# Select by name
map_dbl(x, "x")

# Or by position
map_dbl(x, 1)

# Or by both
map_dbl(x, list("y", 1))

# You'll get an error if a component doesn't exist:
map_chr(x, "z")

# Unless you supply a .default value
map_chr(x, "z", .default = NA)
```

### Passing arguments with ...

It’s often convenient to pass along additional arguments to the function that you’re calling. For example, you might want to pass na.rm = TRUE along to mean(). One way to do that is with an anonymous function:

```{r}
x <- list(1:5, c(1:10, NA))
map_dbl(x, ~ mean(.x, na.rm = TRUE))
```

But because the map functions pass ... along, there’s a simpler form available:

```{r}
map_dbl(x, mean, na.rm = TRUE)
```

This is easiest to understand with a picture: any arguments that come after f in the call to map() are inserted after the data in individual calls to f(): (picture)

Note there’s a subtle difference between placing extra arguments inside an anonymous function compared with passing them to map(). Putting them in an anonymous function means that they will be evaluated every time f() is executed, not just once when you call map(). This is easiest to see if we make the additional argument random:

```{r}
plus <- function(x, y) x + y

x <- c(0, 0, 0, 0)
map_dbl(x, plus, runif(1))

map_dbl(x, ~ plus(.x, runif(1)))
```

### Purrr style


Before we go on to explore more map variants, let’s take a quick look at how you tend to use multiple purrr functions to solve a moderately realistic problem: fitting a model to each subgroup and extracting a coefficient of the model. For this toy example, I’m going to break the mtcars data set down into groups defined by the number of cylinders, using the base split function:

```{r}
by_cyl <- split(mtcars, mtcars$cyl)
```

This creates a list of three data frames: the cars with 4, 6, and 8 cylinders respectively.

Now imagine we want to fit a linear model, then extract the second coefficient (i.e. the slope). The following code shows how you might do that with purrr:

```{r}
by_cyl %>% 
  map(~ lm(mpg ~ wt, data = .x)) %>% 
  map(coef) %>% 
  map_dbl(2)
```

I think this code is easy to read because each line encapsulates a single step, you can easily distinguish the functional from what it does, and the purrr helpers allow us to very concisely describe what to do in each step.

How would you attack this problem with base R? You certainly could replace each purrr function with the equivalent base function:

```{r}
by_cyl %>% 
  lapply(function(data) lm(mpg ~ wt, data = data)) %>% 
  lapply(coef) %>% 
  vapply(function(x) x[[2]], double(1))
```

Or, of course, you could use a for loop:

```{r}
slopes <- double(length(by_cyl))
for (i in seq_along(by_cyl)) {
  model <- lm(mpg ~ wt, data = by_cyl[[i]])
  slopes[[i]] <- coef(model)[[2]]
}
slopes
```

It’s interesting to note that as you move from purrr to base apply functions to for loops you tend to do more and more in each iteration. In purrr we iterate 3 times (map(), map(), map_dbl()), with apply functions we iterate twice (lapply(), vapply()), and with a for loop we iterate once. I prefer more, but simpler, steps because I think it makes the code easier to understand and later modify.

### Map variants

slide

same type of output as input: modify()

Imagine you wanted to double every column in a data frame. You might first try using map(), but map() always returns a list:

```{r}
df <- data.frame(
  x = 1:3,
  y = 6:4
)

map(df, ~ .x * 2)
```

If you want to keep the output as a data frame, you can use modify(), which always returns the same type of output as the input:

```{r}
modify(df, ~ .x * 2)
```

Despite the name, modify() doesn’t modify in place, it returns a modified copy, so if you wanted to permanently modify df, you’d need to assign it:

### Two inputs: map2() and friends

map() is vectorised over a single argument, .x. This means it only varies .x when calling .f, and all other arguments are passed along unchanged, thus making it poorly suited for some problems. For example, how would you find a weighted mean when you have a list of observations and a list of weights? Imagine we have the following data:

```{r}
xs <- map(1:8, ~ runif(10))
xs[[1]][[1]] <- NA
ws <- map(1:8, ~ rpois(10, 5) + 1)
```

You can use map_dbl() to compute the unweighted means:

```{r}
map_dbl(xs, mean)
```

But passing ws as an additional argument doesn’t work because arguments after .f are not transformed:

```{r}
map_dbl(xs, weighted.mean, w = ws)
```

We need a new tool: a map2(), which is vectorised over two arguments. This means both .x and .y are varied in each call to .f:

```{r}
map2_dbl(xs, ws, weighted.mean)
```

picture

The arguments to map2() are slightly different to the arguments to map() as two vectors come before the function, rather than one. Additional arguments still go afterwards:

```{r}
map2_dbl(xs, ws, weighted.mean, na.rm = TRUE)
```

picture

One of the big differences between map2() and the simple function above is that map2() recycles its inputs to make sure that they’re the same length

In other words, map2(x, y, f) will automatically behave like map(x, f, y) when needed. This is helpful when writing functions; in scripts you’d generally just use the simpler form directly.

### No outputs: walk() and friends

Most functions are called for the value that they return, so it makes sense to capture and store the value with a map() function. 

But some functions are called primarily for their side-effects (e.g. cat(), write.csv(), or ggsave()) and it doesn’t make sense to capture their results. 

Take this simple example that displays a welcome message using cat(). cat() returns NULL, so while map() works (in the sense that it generates the desired welcomes), it also returns list(NULL, NULL).

```{r}
welcome <- function(x) {
  cat("Welcome ", x, "!\n", sep = "")
}
names <- c("Hadley", "Jenny")

# As well as generate the welcomes, it also shows 
# the return value of cat()
map(names, welcome)

walk(names, welcome)
```

One of the most useful walk() variants is walk2() because a very common side-effect is saving something to disk, and when saving something to disk you always have a pair of values: the object and the path that you want to save it to.

For example, imagine you have a list of data frames (which was created here using split()), and you’d like to save each one to a separate CSV file. That’s easy with walk2():

```{r}
temp <- tempfile()
dir.create(temp)

cyls <- split(mtcars, mtcars$cyl)
paths <- file.path(temp, paste0("cyl-", names(cyls), ".csv"))
walk2(cyls, paths, write.csv)

dir(temp)
```

### Any number of inputs: pmap() and friends

Since we have map() and map2(), you might expect map3(), map4(), map5(), … But where would you stop? Instead of generalising map2() to an arbitrary number of arguments, purrr takes a slightly different tack with pmap(): you supply it a single list, which contains any number of arguments. In most cases, that will be a list of equal-length vectors, i.e. something very similar to a data frame

```{r}
pmap_dbl(list(xs, ws), weighted.mean, na.rm = TRUE)
```

It’s often convenient to call pmap() with a data frame. 

A handy way to create that data frame is with tibble::tribble(), which allows you to describe a data frame row-by-row (rather than column-by-column, as usual): thinking about the parameters to a function as a data frame is a very powerful pattern. The following example shows how you might draw random uniform numbers with varying parameters:

```{r}
params <- tibble::tribble(
  ~ n, ~ min, ~ max,
   1L,     0,     1,
   2L,    10,   100,
   3L,   100,  1000
)

pmap(params, runif)
```

## Reduce family

After the map family, the next most important family of functions is the reduce family. This family is much smaller, with only two main variants, and is used less commonly, but it’s a powerful idea, gives us the opportunity to discuss some useful algebra, and powers the map-reduce framework frequently used for processing very large datasets.

reduce() takes a vector of length n and produces a vector of length 1 by calling a function with a pair of values at a time: reduce(1:4, f) is equivalent to f(f(f(1, 2), 3), 4).

reduce() is a useful way to generalise a function that works with two inputs (a binary function) to work with any number of inputs. Imagine you have a list of numeric vectors, and you want to find the values that occur in every element. First we generate some sample data:

```{r}
l <- map(1:4, ~ sample(1:10, 15, replace = T))
str(l)
```

To solve this challenge we need to use intersect() repeatedly:

```{r}
out <- l[[1]]
out <- intersect(out, l[[2]])
out <- intersect(out, l[[3]])
out <- intersect(out, l[[4]])
out
```

or

```{r}
reduce(l, intersect)
```

We could apply the same idea if we wanted to list all the elements that appear in at least one entry. All we have to do is switch from intersect() to union()

```{r}
reduce(l, union)
```

Like the map family, you can also pass additional arguments. picture

### Accumulate()

The first reduce() variant, accumulate(), is useful for understanding how reduce works, because instead of returning just the final result, it returns all the intermediate results as well:

```{r}
accumulate(l, intersect)
```

Another useful way to understand reduce is to think about sum(): 

sum(x) is equivalent to x[[1]] + x[[2]] + x[[3]] + ..., i.e. reduce(x, `+`). 

Then accumulate(x, `+`) is the cumulative sum:

```{r}
x <- c(4, 3, 10)
reduce(x, `+`)

accumulate(x, `+`)
```

### Output types

In the above example using +, what should reduce() return when x is short, i.e. length 1 or 0? Without additional arguments, reduce() just returns the input when x is length 1

```{r}
reduce(1, `+`)
```

This means that reduce() has no way to check that the input is valid:

```{r}
reduce("a", `+`)

#but
reduce(integer(), `+`)
```

.init

```{r}
reduce(integer(), `+`, .init = 0)
```

If you want to get algebraic about it, 0 is called the identity of the real numbers under the operation of addition: if you add a 0 to any number, you get the same number back. R applies the same principle to determine what a summary function with a zero length input should return:

```{r}
sum(integer())  # x + 0 = x
#> [1] 0
prod(integer()) # x * 1 = x
#> [1] 1
min(integer())  # min(x, Inf) = x
#> [1] Inf
max(integer())  # max(x, -Inf) = x
#> [1] -Inf
```

If you’re using reduce() in a function, you should always supply .init. Think carefully about what your function should return when you pass a vector of length 0 or 1, and make sure to test your implementation.

### Predicate functionals

A predicate is a function that returns a single TRUE or FALSE, like is.character(), is.null(), or all(), and we say a predicate matches a vector if it returns TRUE.

A predicate functional applies a predicate to each element of a vector. purrr provides seven useful functions which come in three groups:

The following example shows how you might use these functionals with a data frame:

```{r}
df <- data.frame(x = 1:3, y = c("a", "b", "c"))
detect(df, is.factor)

detect_index(df, is.factor)


str(keep(df, is.factor))

str(discard(df, is.factor))
```

### map() variants

map() and modify() come in variants that also take predicate functions, transforming only the elements of .x where .p is TRUE.

```{r}
df <- data.frame(
  num1 = c(0, 10, 20),
  num2 = c(5, 6, 7),
  chr1 = c("a", "b", "c"),
  stringsAsFactors = FALSE
)

str(map_if(df, is.numeric, mean))

str(modify_if(df, is.numeric, mean))

str(map(keep(df, is.numeric), mean))
```

### Base mathematical functionals

integrate() finds the area under the curve defined by f()
uniroot() finds where f() hits zero
optimise() finds the location of the lowest (or highest) value of f()

```{r}
integrate(sin, 0, pi)

str(uniroot(sin, pi * c(1 / 2, 3 / 2)))

str(optimise(sin, c(0, 2 * pi)))

str(optimise(sin, c(0, pi), maximum = TRUE))
```

