---
title: "R functions"
---
Some very interesting questions:

1. What are the three components of a function?

2. What does the following code return?

```{r}
x <- 10
f1 <- function(x) {
  function() {
    x + 10
  }
}

f1(1)()
```

3. How would you usually write this code?

```{r}
`+`(1, `*`(2, 3))
```

4. How could you make this call easier to read?

```{r}
mean(, TRUE, x = c(1:10, NA))
```

5. Does the following code throw an error when executed? Why or why not?

```{r}
f2 <- function(a, b) {
  a * 10
}
f2(10, stop("This is an error!"))
```

7. How do you ensure that cleanup action occurs regardless of how a function exits?


A function has three parts:

The **formals()**, the list of arguments that control how you call the function.

The **body()**, the code inside the function.

The **environment()**, the data structure that determines how the function finds the values associated with the names.

While the formals and body are specified explicitly when you create a function, the environment is specified implicitly, based on where you defined the function. The function environment always exists, but it is only printed when the function isn’t defined in the global environment.

```{r}
foo1 <- function(x = 0, y = 1, z) {c(x,y,z)}
environment(foo1)
```


```{r}
f02 <- function(x, y) {
  # A comment
  x + y
}

formals(f02)

body(f02)

environment(f02)
```

```{r}
attributes(f02)

attr(f02, "srcref")
```

### First-class functions

While you almost always create a function and then bind it to a name, the binding step is not compulsory. If you choose not to give a function a name, you get an anonymous function. This is useful when it’s not worth the effort to figure out a name:

```{r}
mtcars

lapply(mtcars, function(x) length(unique(x)))

Filter(function(x) is.numeric(x), mtcars)
```

A final option is to put functions in a list:

```{r}
funs <- list(
  half = function(x) x / 2,
  double = function(x) x * 2
)

funs$half(10)
```

```{r}
funs$data <- c('my','data')
funs
```

### Invoking functions

You normally call a function by placing its arguments, wrapped in parentheses, after its name: mean(1:10, na.rm = TRUE). 

But what happens if you have the arguments already in a data structure?

```{r}
args <- list(1:10, na.rm = TRUE)
```

You can instead use do.call(): it has two arguments. The function to call, and a list containing the function arguments:

```{r}
do.call(mean, args)
mean(1:10, na.rm = TRUE)
```

### Function composition

Base R provides two ways to compose multiple function calls. For example, imagine you want to compute the population standard deviation using sqrt() and mean() as building blocks:

```{r}
square <- function(x) x^2
deviation <- function(x) x - mean(x)
```

You either nest the function calls:

```{r}
x <- runif(100)

sqrt(mean(square(deviation(x))))
```

Or you save the intermediate results as variables:

```{r}
out <- deviation(x)
out <- square(out)
out <- mean(out)
out <- sqrt(out)
out
```


The magrittr package provides a third option: 

the binary operator %>%, which is called the pipe and is pronounced as “and then”.

```{r}
library(magrittr)
```

```{r}
x %>%
  deviation() %>%
  square() %>%
  mean() %>%
  sqrt()
```

x %>% f() is equivalent to f(x); x %>% f(y) is equivalent to f(x, y). 

The pipe allows you to focus on the high-level composition of functions rather than the low-level flow of data; 

the focus is on what’s being done (the verbs), rather than on what’s being modified (the nouns). This style is common in Haskell and F#, the main inspiration for magrittr.

```{r}
iris

iris %>% head() %>% summary()

summary(head(iris))
```


### Scoping

1. Name masking

The basic principle of lexical scoping is that names defined inside a function mask names defined outside a function. This is illustrated in the following example.


```{r}
x <- 10
y <- 20
g02 <- function() {
  x <- 1
  y <- 2
  c(x, y)
}
g02()
```

If a name isn’t defined inside a function, R looks one level up.

```{r}
x <- 2
g03 <- function() {
  y <- 1
  c(x, y)
}
g03()

# And this doesn't change the previous value of y
y
```

The same rules apply if a function is defined inside another function. First, R looks inside the current function. Then, it looks where that function was defined (and so on, all the way up to the global environment). Finally, it looks in other loaded packages.

```{r}
x <- 1
g04 <- function() {
  y <- 2
  i <- function() {
    z <- 3
    c(x, y, z)
  }
  i()
}

g04()
```

2. functions vs variables

In R, functions are ordinary objects. This means the scoping rules described above also apply to functions:

```{r}
g07 <- function(x) x + 1

g08 <- function() {
  g07 <- function(x) x + 100
  g07(10)
}

g08()
```

However, when a function and a non-function share the same name (they must, of course, reside in different environments), applying these rules gets a little more complicated. When you use a name in a function call, R ignores non-function objects when looking for that value. For example, in the code below, g09 takes on two different values:

```{r}
g09 <- function(x) x + 100

g10 <- function() {
  g09 <- 10
  g09(g09)
}

g10()
```

3. A fresh start

```{r}
g11 <- function() {
  if (!exists("a")) {
    a <- 1
  } else {
    a <- a + 1
  }
  a
}

g11()
g11()
```

4. Dynamic lookup

Lexical scoping determines where, but not when to look for values. R looks for values when the function is run, not when the function is created. Together, these two properties tell us that the output of a function can differ depending on the objects outside the function’s environment:

```{r}
g12 <- function() x + 1
x <- 15
g12()

x <- 20
g12()
```

This behaviour can be quite annoying. If you make a spelling mistake in your code, you won’t get an error message when you create the function. And depending on the variables defined in the global environment, you might not even get an error message when you run the function.

To detect this problem, use codetools::findGlobals(). This function lists all the external dependencies (unbound symbols) within a function:

```{r}
codetools::findGlobals(g12)
```

```{r}
environment(g09)
environment(g12) <- emptyenv()
g12()
```

The problem and its solution reveal why this seemingly undesirable behaviour exists: R relies on lexical scoping to find everything, from the obvious, like mean(), to the less obvious, like + or even {. This gives R’s scoping rules a rather beautiful simplicity.

Ex
```{r}
c <- 10
c(c = c)
```


### Lazy evaluation

Default arguments

Thanks to lazy evaluation, default values can be defined in terms of other arguments, or even in terms of variables defined later in the function:

```{r}
h04 <- function(x = 1, y = x * 2, z = a + b) {
  a <- 10
  b <- 100
  
  c(x, y, z)
}

h04()
```


The evaluation environment is slightly different for default and user supplied arguments, as default arguments are evaluated inside the function. This means that seemingly identical calls can yield different results. It’s easiest to see this with an extreme example:

```{r}
h05 <- function(x = ls()) {
  a <- 1
  x
}

# ls() evaluated inside h05:
h05()

# ls() evaluated in global environment:
h05(ls())
ls()
```




...

```{r}
foo <- function(...) c(..1, ..4)
foo(1,2,3,'a')



foo <- function(...) {
  list(..1, ..2)
  }

foo(x=20, y=30, z=500)



foo <- function(...) {
  list(...)
  }

foo(x=20, y=30, z=500, a = 'another')
```

```{r}
sum(1, 2, 3)
#> [1] 6
mean(7, 2, 3)
#> [1] 1
mean(c(1,2,3))

sum(1, 2, 3, na.omit = TRUE)
#> [1] 7
mean(1, 2, 3, na.omit = TRUE)
#> [1] 1


sum(TRUE, 1, 2, 0.5)
?mean
```

```{r}
plot(1:10, col = "red", pch = 10, xlab = "x", col.lab = "blue")
```

```{r}
?plot
```

